Mapping the current knowledge in syndemic research applied to Men who have Sex with Men : a Scoping Review

### 2022-03-31 _bookdown_files/02-introduction.Rmd
References correction

Men who have sex with men (MSM) represent a vulnerable population disproportionately affected by numerous health conditions. They represent 70% of new HIV diagnoses @centerfordiseasecontrolandprevention2020 in the United States of America and more than half of new HIV diagnoses in European Union/European Economic area @europeancentrefordiseasepreventionandcontrol2019. They are also more prone to contract other sexually transmitted diseases and *Nesseria Gonorrhea* strains from MSM exhibit higher antimicrobial resistance @centersfordiseasecontrolandprevention2019. Furthermore, mental health conditions such as depression, anxiety, suicide attempts or self-harm are more prevalent among MSM [@ross2018; @Liu2019; @luo2017] and substance use is more common than for their heterosexual counterparts @Medley2016. Moreover, those adverse health outcomes are enmeshed within structural disadvantages such as violence, stigma, discrimination [@collier2013; @Lee2016; @Lea2014] as well as with poverty, unemployment, unstable housing and poor healthcare access [@closson2018; @ayhan2020].

For more than 20 years, syndemic theory has provided a framework to examine the interrelations between social conditions, mental health and physical health @singer2017. Described for the first time in 1996 by Merrill Singer @singer1996, a syndemic consists of two more interacting epidemics producing an excess burden in a population because of harmful social conditions @singer2003. To speak of a syndemic, three conditions must be met @mendenhall2020 :

@@

Syndemic is thus an holistic framework, describing interactions both between diseases themselves and between diseases and the social environment contributing to their emergence, clustering and spread @singer2017. As such, its value for studying the health of marginalized populations such as MSM is undeniable.

@@

However, this theory is not without its flaws. A systematic review conducted in 2015 showed the inadequacy of most of the literature in supporting one of its core tenets @tsai2015. Indeed, this paper demonstrated that, although synergistic interaction between diseases lies in the core of the theory, most papers failed to use relevant statistics to demonstrate the existence of a synergy. More recently, a scoping review published in 2020 confirmed this finding as most recently-published citations failed to describe the interactions between diseases needed to account for a true syndemic @Singer2020. Scholars in the field have been calling for greater clarity in the use of the concept and urged researchers to describe more precisely the interactions they observe between health conditions [@singer2021 ; @mendenhall2020].

>>>>

Men who have sex with men (MSM) represent a vulnerable population disproportionately affected by numerous health conditions. They represent 70% of new HIV diagnoses [@centerfordiseasecontrolandprevention2020] in the United States of America and more than half of new HIV diagnoses in European Union/European Economic area [@europeancentrefordiseasepreventionandcontrol2019]. They are also more prone to contract other sexually transmitted diseases and *Nesseria Gonorrhea* strains from MSM exhibit higher antimicrobial resistance [@centersfordiseasecontrolandprevention2019]. Furthermore, mental health conditions such as depression, anxiety, suicide attempts or self-harm are more prevalent among MSM [@ross2018; @Liu2019; @luo2017] and substance use is more common than for their heterosexual counterparts[@Medley2016]. Moreover, those adverse health outcomes are enmeshed within structural disadvantages such as violence, stigma, discrimination [@collier2013; @Lee2016; @Lea2014] as well as with poverty, unemployment, unstable housing and poor healthcare access [@closson2018; @ayhan2020].

For more than 20 years, syndemic theory has provided a framework to examine the interrelations between social conditions, mental health and physical health [@singer2017]. Described for the first time in 1996 by Merrill Singer [@singer1996], a syndemic consists of two more interacting epidemics producing an excess burden in a population because of harmful social conditions [@singer2003]To speak of a syndemic, three conditions must be met [@mendenhall2020] :

@@

Syndemic is thus an holistic framework, describing interactions both between diseases themselves and between diseases and the social environment contributing to their emergence, clustering and spread [@singer2017]. As such, its value for studying the health of marginalized populations such as MSM is undeniable.

@@

However this framework is frequently misused by researchers. A systematic review conducted in 2015 showed the inadequacy of most of the literature in supporting one of its core tenets [@tsai2015]. Indeed, this paper demonstrated that, although synergistic interaction between diseases lies in the core of the theory, most papers failed to use relevant statistics to demonstrate the existence of a synergy. More recently, a scoping review published in 2020 confirmed this finding as most recently-published citations failed to describe the interactions between diseases needed to account for a true syndemic [@Singer2020]. Scholars in the field have been calling for greater clarity in the use of the concept and urged researchers to describe more precisely the interactions they observe between health conditions [@singer2021 ; @mendenhall2020].

### 2022-03-31 _bookdown_files/03-methodology.Rmd
References correction

We applied the framework suggested by Arskey and O'Malley @arksey2005, with enhancements from Levac et al. @levac2010. We also took into account the recommendations formulated by Colqhoun et al. @colquhoun2014 and the updated guidance by Peters and colleagues @peters2020. We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses extension for Scoping Reviews (PRISMA-ScR) @Tricco2018. The completed checklist can be found in \@ref(tab:PRISMA-ScR) of the supplementary materials.

Screening and data charting was completed using DistillerSR (Evidence Partners, Ottawa, Canada). Scripts, statistics, charts, tables and the present paper was generated using R @rcoreteam2020, RStudio @rstudioteam2020 and appropriate packages. To ensure the complete transparency and reproducibility of this article, our data manager (M.O.) uploaded every data used to generate this article on Open Science Framework (DOI : [masked for the anonymity of the review])

>>>>

We applied the framework suggested by Arskey and O'Malley [@arksey2005], with enhancements from Levac et al. [@levac2010]. We also took into account the recommendations formulated by Colqhoun et al. [@colquhoun2014] and the updated guidance by Peters and colleagues [@peters2020]. We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses extension for Scoping Reviews (PRISMA-ScR) [@Tricco2018]. The completed checklist can be found in \@ref(tab:PRISMA-ScR) of the supplementary materials.

Screening and data charting was completed using DistillerSR (Evidence Partners, Ottawa, Canada). Scripts, statistics, charts, tables and the present paper was generated using R [@rcoreteam2020], RStudio [@rstudioteam2020] and appropriate packages. To ensure the complete transparency and reproducibility of this article, our data manager (M.O.) uploaded every data used to generate this article on Open Science Framework (DOI : [masked for the anonymity of the review])

### 2022-03-31 _bookdown_files/04-results.Rmd
References correction

The electronic search identified 840 references and our handsearch of reference lists resulted in the addition of the seminal study by Stall and colleague @Stall2003 . After removing duplicates, 305 records were screened for inclusion.

>>>>

The electronic search identified 840 references and our handsearch of reference lists resulted in the addition of the seminal study by Stall and colleague [@Stall2003]. After removing duplicates, 305 records were screened for inclusion.

### 2022-03-31 _bookdown_files/05-discussion.Rmd
References correction

Three issues arise from this. First, under-representation of the Global South compromises our understanding of the health of MSM living in these countries. Furthermore, as pointed out by @weaver2020, the screening tool used to assess the presence of a syndemic condition in the Western context may not be efficient in another sociocultural setting, especially in regard to mental health. As syndemics are driven by the sociocultural forces of a given place at a given time, these potential biases need to be addressed in order to conduct meaningful research outside Europe and North America.

Moreover, even for studies conducted in Europe and North America, a large portion of the studies conflate data from different cities, counties, states or even countries in order to obtain a sample big enough to conduct statistical analyses. However, geographical variations in syndemic burden are likely to be missed. Migration of MSM to larger cities is a well-known phenomenon and thought to be an important part of syndemic production among MSM as well as a source of resilience [@stall2007; @Bruce2011]. Differences in the density of gay social venues and peer organizations as well as exposure to discrimination and violence or an easier access to substances need to be taken into account when studying a syndemic. As @shrestha2020 suggested, syndemic research needs to incorporate data from spatial epidemiology to better understand the interactions of socioeconomic forces in the physical space and how these forces can influence the health of MSM.

@@

The design of syndemic studies is another element suffering from a lack of diversity. The field is dominated by quantitative studies as qualitative papers represent only `r percent(.N_qual/nrow(overview_of_studies))` of our sample. Moreover `r percent(.N_cross_sec/nrow(summary_quantitative))` of the quantitative studies were cross-sectional. As vulnerability to syndemic conditions is suspected to arise from early adverse experience @stall2007 longitudinal studies running over a long period of time are probably needed to study syndemic production among MSM.

Moreover, as shown in our result, qualitative papers are more prone to discuss the potential interactions between syndemic conditions. Some concepts and experiences needed to understand syndemic production in MSM can only be studied through a qualitative approach. For example, the complex narratives surrounding drug use and its link to loneliness, marginalization, personal affirmation, resistance and lack of meaningful emotional connections @Pollard2018a would be hard to measure and model using quantitative data. As Tsai previously pointed out, the inevitable simplification needed to estimate statistical models may lead to an over-simplification of diseases dynamics @Tsai2018a . Mixed method are needed to complement robust statistical estimation with insights on sociocultural mechanisms and on the lived experiences of people facing this clustering of adverse conditions. Unfortunately, our review only identified `r max(year_design$cumulative[year_design$design == "Mixed"])` studies employing such mixed methodology. Future research would benefit from generating and analysing qualitative data to support and enrich their quantitative findings.

Finally, we identified `r .N_int` interventional studies with mixed results. While co-targeting interconnected syndemic conditions is supported by a meta-analysis @Pantalone2020, a better understanding of how syndemics operate may be needed before we are able to offer efficient syndemic-based interventions.

@@

Studying the health of MSM poses a challenge regarding the diversity of this population. While nearly half of the studies included in our sample studied a subpopulation of MSM, the number of MSM subpopulation is such that most MSM subpopulation remains under-served. For example, Young Black MSM are specifically studied in `r .N_YBMSM` studies while representing 52% of the new HIV diagnosis among MSM aged 13-24 years in 2018 @centerfordiseasecontrolandprevention2020.

This discrepancy between the number of studies focusing on a subpopulation and the very high prevalence of some syndemic condition such as HIV infection is even more concerning when considering that dual-minority identity may expose a person to different stressors and outcomes. For example, some studies found that Black MSM had a higher rate of HIV seroconversion than White MSM while being exposed to fewer syndemic conditions @Mustanski2017 . These findings are supported by qualitative data suggesting that intersectional theory offers the context needed to understand syndemics among Black MSM which differ from syndemics among White MSM @Quinn. Furthermore, the findings from Bulled and colleagues suggest that structural disadvantage may have more weight on the lives of Black MSM than individual factors such as substance use @bulled2021

MSMW are another example of an under-served population with conflicting results. In a cross-sectional study using syndemic sum count, Latino MSMW with two or more syndemic conditions had a 7.09 Odds Ratio to engage in receptive condomless anal sex but no statistically significant difference was found for insertive condomless anal sex or condomless vaginal sex @Munoz-Laboy2018. On the other hand, in another study the number of syndemic conditions was associated with HIV diagnosis for MSM but not for MSW or MSMW @Branstrom2018. Among young MSM and MSMW, one study found more syndemic conditions and a greater association between them for MSMW compared to MSM @Mustanski2014. On the contrary, Ferlatte et al. found that MSM were more likely than MSMW to suffer from two or more syndemic conditions @ferlatte2018. These conflicting findings may be partly due to the difficulty to define and operate bisexuality in research @swan2018, differences between age groups or differences in syndemic mechanism for MSMW compared to MSM. Outside the field of syndemic literature, studies suggest that bisexual men suffer from a similar or higher burden of mental health conditions compared to gay men [@salway2019; @Chaudhry2019; @ross2018] and literature specific to bisexual men remains scarce. In the future, syndemic theory should thus seek to study the specificity of this population. First, there's a need to better conceptualize and define bisexuality in the field - should we still use solely behavior-based definition (as in MSMW) or should we incorporate a self-definition of one's sexual orientation ? Then, researchers should aim to disaggregate data concerning MSMW/bisexual men from data concerning MSM/gay men whenever possible. Lastly, more research focusing on MSMW/bisexual men is needed.

Finally transgender MSM were included in only one study in our sample. As stated in our results, this is concerning, especially when every other study excludes transgender men by design. As such, this population of MSM is the one we know the least about. The study from Reisner and colleagues suggested that transgender MSM who had socially affirmed their gender were exposed to similar risk patterns as cisgender MSM @Reisner2016a. The health of transgender men remains greatly understudied and the few available studies suggest high rates of adverse conditions such as violence, notably sexual violence @Testa2012, suicide attempts @Haas2014, chlamydial and gonoccocal infections @Pitasi2019, intimate partner violence [@peitzmeier2020] and HIV infection @Becasen2019. Researchers should be encouraged to include transgender men in their studies and to conduct research focused on transgender MSM. Indeed, syndemic theory represents a promising framework for research on transgender men as its premises of social adversities producing and concentrating diseases in a population allows us to explore the health of transgender men without pathologizing them.

@@

This omnipresence of HIV in syndemic literature applied to MSM is understandable. The first description of a syndemic, the SAVA syndemic, incorporated AIDS as part of a syndemic @singer1996. Then, a few years later, the first paper on syndemic in the MSM population had HIV-related outcomes, in the form of HIV serostatus and CAS @Stall2003. Furthermore MSM were the first and most severely affected population when the AIDS epidemic started @decock2012. Nowadays, male-to-male sexual contact still represents nearly 70% of the new HIV cases in the USA @centerfordiseasecontrolandprevention2020. While we can't deny the importance of the HIV pandemic, notably for MSM, we would argue that HIV is not the only issue in need of attention.

Indeed, MSM are also disproportionately affected by, amongst others, suicide, anxiety, depression and substance use[@ross2018; @luo2017; @Medley2016]. Moreover, with the advances in antiretroviral therapy and the subsequent decline in HIV-related mortality, a Canadian study showed that gay men were more likely to die from suicide than from HIV in 2011 @hottes2015. Even if the estimates Hottes' paper overestimated the mortality due to suicide among MSM, suicide would still be a leading cause of death in this population while being studied as an outcome in only `r sum(!is.na(.outcomes$outcome_suicide))` papers in our sample.

Furthermore, when studying HIV as an outcome, we would argue that using condomless anal sex as a proxy should be reviewed and better proxies sought. Indeed, condomless anal sex was one of the criteria in `r .CAS_number` studies out of the `r sum(!is.na(.outcomes$outcome_sexual_risk))` with sexual risk behaviors as an outcome and `r sum(!is.na(.outcomes_sex_risk_measurement$outcome_CAS))` studies considered all form of condomless anal sex as a risk behavior. However, without refinement, this outcome fails to capture the changes in HIV prevention strategies such as PrEP or Treatment as Prevention (TasP). Only `r english(sum(!is.na(.outcomes_sex_risk_measurement$outcome_PrEP_VL_CAS)))` studies in our sample took PrEP use and an undetectable viral load into account when considering condomless anal sex as a risk factor for HIV acquisition. When the first paper on syndemic theory applied to MSM was published, engaging in condomless anal sex would have led to a high risk of acquiring HIV. Nowadays, people could be on PrEP, forego the use of a condom and still be at a very low risk of acquiring HIV @calabrese2017 and our methodologies in syndemic research should reflect this evolution. In addition, qualitative research gives an insight into risk behaviors and offers a more nuanced view on this topic than a binary variable of consistent condom use for anal intercourse. One such paper published in 2017 showed that risk practices fell into different subjectivities such as active and consistent pursuit of condomless sex, lack of assertiveness to a partner's initiative of condomless anal sex and combination of episodic risk practices and reduction strategies @Adam2017. Besides, some studies showed that participants with a greater number of syndemic conditions would engage more frequently in condomless anal sex while not being more likely to test positive for HIV [@Mustanski2017; @Pitpitan2016]. On the other hand, a paper from 2018 found that having a greater number of syndemic conditions was associated to self-reported HIV but not to condomless anal sex @chuang2018. Taken together, we thus postulate that condomless anal sex should at least be replaced by "condomless anal sex without consistent PrEP use or with a detectable viral load" when an investigator seek to use sexual risk behaviors as an outcome.

@@

One of the tenets of syndemic theory is that there exists some form of biological, social and/or psychological interaction between the diseases @mendenhall2020. As others have already pointed out, the extent to which syndemic literature had proved this core principle remains very limited @tsai2015. Indeed, `r percent(.score_data/nrow(summary_quantitative))` of our quantitative sample of studies used solely a summation score of the syndemic conditions to conduct regression analysis. Two issues arise from this statistical approach. First, summation score are additive by essence, thus unsuitable to demonstrate an interaction @tsai2015. Second, two psychometric assumptions are made by such a model : (1) the unidimensionality of the construct and (2) equal factor weighting @Halkitis2013a. Put simply, using a summation score would imply that (1) every condition forms a single construct and that (2) every condition contributes equally to said construct (which would mean for example that feeling depressed during the past 14 days and having an history of childhood sexual abuse would contribute equally to a syndemic). The unidimensionality aspect of the construct has received some empirical support by Mustanski et al. as well as by Starks et al. using respectively Structural Equation Modeling and Latent Class Analysis [@Mustanski2014; @Starks2014]. On the contrary, Leblanc et al. failed to construct a single latent syndemic variable in their analysis [@leblanc2021]. Moreover constraining factor loadings to be equal resulted in worse fit in the studies by Mustanski and Starks.

@@

More recently, a new statistical approach inspired from research in psychopathology has been used to model syndemics : network analysis [@Lee; @lee2020]. The core principle of this approach is that mental disorders arise from causal interactions between symptoms in a network @borsboom2017. Applied to syndemic theory, a syndemic would then be the network and the various conditions would then be the nodes of the network. Furthermore, one of the benefits of this approach resides in the theoretical possibility to identify influential nodes in a network : nodes that play a significant part in the activation of the network and that could be the prime target of an intervention in order to deactivate the network @robinaugh2016. In short, in addition to being statistically sound and coherent with the model of mutually causal epidemics, network analysis could offer actionable data to improve the health of MSM by identifying which conditions should be prioritized for interventions [@Tsai2018a].

We recommend that future research should abandon the summation score approach and attempt to clarify which model of interaction receives the most empirical support in order to implement effective public health interventions. Indeed as reminded by Chakrapani et al. the three models have different programmatic implications @Chakrapani2019. In the case of synergistically interacting epidemics, a single intervention may yield greater health improvement than if no interaction was present. For mutually causal epidemics, multicomponent interventions addressing the various epidemics should be implemented in order to produce an improvement. Finally, for serially causal epidemics, intervening at the root cause of a syndemic may prevent the development of a syndemic cascade.

@@

In qualitative studies, resilience was also found to exert a protective effect against syndemic conditions [@Chakrapani2019a; @Reed2016; @adam2018]. Furthermore, the review by Woodwards et al. found a protective effect of most resilience resources identified against HIV @woodward2017.

@@

Minority stress represents another interesting framework as it allows to explain how social stigma can get under the skin of minorized populations @Meyer2003.

@@

Some scholars are skeptical on the usefulness of syndemic theory to intersectionality and question the possibility of a conversation between the two fields @sangaramoorthy2021. As pertinent as the authors commentary is, we do believe in the possibility of making the two field converse. As Quinn pointed out in her response, incorporating an intersectional framework into syndemic research may address issues of power, oppression and structural violence, neglected thus far in the field @quinn2021. Doing so, we may attend to some gap in the literature and disentangle the complex web of stigma and structural violence faced by sexual minority men, especially MSM situated at the intersection of multiple form of systemic discrimination such as racism, homophobia, biphobia, transphobia, classism, ableism or stigma against sex workers or people living with HIV.

>>>>

Three issues arise from this. First, under-representation of the Global South compromises our understanding of the health of MSM living in these countries. Furthermore, as pointed out by [@weaver2020], the screening tool used to assess the presence of a syndemic condition in the Western context may not be efficient in another sociocultural setting, especially in regard to mental health. As syndemics are driven by the sociocultural forces of a given place at a given time, these potential biases need to be addressed in order to conduct meaningful research outside Europe and North America.

Moreover, even for studies conducted in Europe and North America, a large portion of the studies conflate data from different cities, counties, states or even countries in order to obtain a sample big enough to conduct statistical analyses. However, geographical variations in syndemic burden are likely to be missed. Migration of MSM to larger cities is a well-known phenomenon and thought to be an important part of syndemic production among MSM as well as a source of resilience [@stall2007; @Bruce2011]. Differences in the density of gay social venues and peer organizations as well as exposure to discrimination and violence or an easier access to substances need to be taken into account when studying a syndemic. As [@shrestha2020] suggested, syndemic research needs to incorporate data from spatial epidemiology to better understand the interactions of socioeconomic forces in the physical space and how these forces can influence the health of MSM.

@@

The design of syndemic studies is another element suffering from a lack of diversity. The field is dominated by quantitative studies as qualitative papers represent only `r percent(.N_qual/nrow(overview_of_studies))` of our sample. Moreover `r percent(.N_cross_sec/nrow(summary_quantitative))` of the quantitative studies were cross-sectional. As vulnerability to syndemic conditions is suspected to arise from early adverse experience [@stall2007] longitudinal studies running over a long period of time are probably needed to study syndemic production among MSM.

Moreover, as shown in our result, qualitative papers are more prone to discuss the potential interactions between syndemic conditions. Some concepts and experiences needed to understand syndemic production in MSM can only be studied through a qualitative approach. For example, the complex narratives surrounding drug use and its link to loneliness, marginalization, personal affirmation, resistance and lack of meaningful emotional connections [@Pollard2018a] would be hard to measure and model using quantitative data. As Tsai previously pointed out, the inevitable simplification needed to estimate statistical models may lead to an over-simplification of diseases dynamics [@Tsai2018a]. Mixed method are needed to complement robust statistical estimation with insights on sociocultural mechanisms and on the lived experiences of people facing this clustering of adverse conditions. Unfortunately, our review only identified `r max(year_design$cumulative[year_design$design == "Mixed"])` studies employing such mixed methodology. Future research would benefit from generating and analysing qualitative data to support and enrich their quantitative findings.

Finally, we identified `r .N_int` interventional studies with mixed results. While co-targeting interconnected syndemic conditions is supported by a meta-analysis [@Pantalone2020], a better understanding of how syndemics operate may be needed before we are able to offer efficient syndemic-based interventions.

@@

Studying the health of MSM poses a challenge regarding the diversity of this population. While nearly half of the studies included in our sample studied a subpopulation of MSM, the number of MSM subpopulation is such that most MSM subpopulation remains under-served. For example, Young Black MSM are specifically studied in `r .N_YBMSM` studies while representing 52% of the new HIV diagnosis among MSM aged 13-24 years in 2018 [@centerfordiseasecontrolandprevention2020].

This discrepancy between the number of studies focusing on a subpopulation and the very high prevalence of some syndemic condition such as HIV infection is even more concerning when considering that dual-minority identity may expose a person to different stressors and outcomes. For example, some studies found that Black MSM had a higher rate of HIV seroconversion than White MSM while being exposed to fewer syndemic conditions [@Mustanski2017] . These findings are supported by qualitative data suggesting that intersectional theory offers the context needed to understand syndemics among Black MSM which differ from syndemics among White MSM [@Quinn]. Furthermore, the findings from Bulled and colleagues suggest that structural disadvantage may have more weight on the lives of Black MSM than individual factors such as substance use [@bulled2021]

MSMW are another example of an under-served population with conflicting results. In a cross-sectional study using syndemic sum count, Latino MSMW with two or more syndemic conditions had a 7.09 Odds Ratio to engage in receptive condomless anal sex but no statistically significant difference was found for insertive condomless anal sex or condomless vaginal sex [@Munoz-Laboy2018]. On the other hand, in another study the number of syndemic conditions was associated with HIV diagnosis for MSM but not for MSW or MSMW [@Branstrom2018]. Among young MSM and MSMW, one study found more syndemic conditions and a greater association between them for MSMW compared to MSM [@Mustanski2014]. On the contrary, Ferlatte et al. found that MSM were more likely than MSMW to suffer from two or more syndemic conditions [@ferlatte2018]. These conflicting findings may be partly due to the difficulty to define and operate bisexuality in research [@swan2018], differences between age groups or differences in syndemic mechanism for MSMW compared to MSM. Outside the field of syndemic literature, studies suggest that bisexual men suffer from a similar or higher burden of mental health conditions compared to gay men [@salway2019; @Chaudhry2019; @ross2018] and literature specific to bisexual men remains scarce. In the future, syndemic theory should thus seek to study the specificity of this population. First, there's a need to better conceptualize and define bisexuality in the field - should we still use solely behavior-based definition (as in MSMW) or should we incorporate a self-definition of one's sexual orientation ? Then, researchers should aim to disaggregate data concerning MSMW/bisexual men from data concerning MSM/gay men whenever possible. Lastly, more research focusing on MSMW/bisexual men is needed.

Finally transgender MSM were included in only one study in our sample. As stated in our results, this is concerning, especially when every other study excludes transgender men by design. As such, this population of MSM is the one we know the least about. The study from Reisner and colleagues suggested that transgender MSM who had socially affirmed their gender were exposed to similar risk patterns as cisgender MSM [@Reisner2016a]. The health of transgender men remains greatly understudied and the few available studies suggest high rates of adverse conditions such as violence, notably sexual violence [@Testa2012], suicide attempts [@Haas2014], chlamydial and gonoccocal infections [@Pitasi2019], intimate partner violence [@peitzmeier2020] and HIV infection [@Becasen2019]. Researchers should be encouraged to include transgender men in their studies and to conduct research focused on transgender MSM. Indeed, syndemic theory represents a promising framework for research on transgender men as its premises of social adversities producing and concentrating diseases in a population allows us to explore the health of transgender men without pathologizing them.

@@

This omnipresence of HIV in syndemic literature applied to MSM is understandable. The first description of a syndemic, the SAVA syndemic, incorporated AIDS as part of a syndemic [@singer1996]. Then, a few years later, the first paper on syndemic in the MSM population had HIV-related outcomes, in the form of HIV serostatus and CAS [@Stall2003]. Furthermore MSM were the first and most severely affected population when the AIDS epidemic started [@decock2012]. Nowadays, male-to-male sexual contact still represents nearly 70% of the new HIV cases in the USA [@centerfordiseasecontrolandprevention2020]. While we can't deny the importance of the HIV pandemic, notably for MSM, we would argue that HIV is not the only issue in need of attention.

Indeed, MSM are also disproportionately affected by, amongst others, suicide, anxiety, depression and substance use[@ross2018; @luo2017; @Medley2016]. Moreover, with the advances in antiretroviral therapy and the subsequent decline in HIV-related mortality, a Canadian study showed that gay men were more likely to die from suicide than from HIV in 2011 [@hottes2015]. Even if the estimates Hottes' paper overestimated the mortality due to suicide among MSM, suicide would still be a leading cause of death in this population while being studied as an outcome in only `r sum(!is.na(.outcomes$outcome_suicide))` papers in our sample.

Furthermore, when studying HIV as an outcome, we would argue that using condomless anal sex as a proxy should be reviewed and better proxies sought. Indeed, condomless anal sex was one of the criteria in `r .CAS_number` studies out of the `r sum(!is.na(.outcomes$outcome_sexual_risk))` with sexual risk behaviors as an outcome and `r sum(!is.na(.outcomes_sex_risk_measurement$outcome_CAS))` studies considered all form of condomless anal sex as a risk behavior. However, without refinement, this outcome fails to capture the changes in HIV prevention strategies such as PrEP or Treatment as Prevention (TasP). Only `r english(sum(!is.na(.outcomes_sex_risk_measurement$outcome_PrEP_VL_CAS)))` studies in our sample took PrEP use and an undetectable viral load into account when considering condomless anal sex as a risk factor for HIV acquisition. When the first paper on syndemic theory applied to MSM was published, engaging in condomless anal sex would have led to a high risk of acquiring HIV. Nowadays, people could be on PrEP, forego the use of a condom and still be at a very low risk of acquiring HIV [@calabrese2017] and our methodologies in syndemic research should reflect this evolution. In addition, qualitative research gives an insight into risk behaviors and offers a more nuanced view on this topic than a binary variable of consistent condom use for anal intercourse. One such paper published in 2017 showed that risk practices fell into different subjectivities such as active and consistent pursuit of condomless sex, lack of assertiveness to a partner's initiative of condomless anal sex and combination of episodic risk practices and reduction strategies [@Adam2017]. Besides, some studies showed that participants with a greater number of syndemic conditions would engage more frequently in condomless anal sex while not being more likely to test positive for HIV [@Mustanski2017; @Pitpitan2016]. On the other hand, a paper from 2018 found that having a greater number of syndemic conditions was associated to self-reported HIV but not to condomless anal sex [@chuang2018]. Taken together, we thus postulate that condomless anal sex should at least be replaced by "condomless anal sex without consistent PrEP use or with a detectable viral load" when an investigator seek to use sexual risk behaviors as an outcome.

@@

One of the tenets of syndemic theory is that there exists some form of biological, social and/or psychological interaction between the diseases [@mendenhall2020]. As others have already pointed out, the extent to which syndemic literature had proved this core principle remains very limited [@tsai2015]. Indeed, `r percent(.score_data/nrow(summary_quantitative))` of our quantitative sample of studies used solely a summation score of the syndemic conditions to conduct regression analysis. Two issues arise from this statistical approach. First, summation score are additive by essence, thus unsuitable to demonstrate an interaction [@tsai2015]. Second, two psychometric assumptions are made by such a model : (1) the unidimensionality of the construct and (2) equal factor weighting [@Halkitis2013a]. Put simply, using a summation score would imply that (1) every condition forms a single construct and that (2) every condition contributes equally to said construct (which would mean for example that feeling depressed during the past 14 days and having an history of childhood sexual abuse would contribute equally to a syndemic). The unidimensionality aspect of the construct has received some empirical support by Mustanski et al. as well as by Starks et al. using respectively Structural Equation Modeling and Latent Class Analysis [@Mustanski2014; @Starks2014]. On the contrary, Leblanc et al. failed to construct a single latent syndemic variable in their analysis [@leblanc2021]. Moreover constraining factor loadings to be equal resulted in worse fit in the studies by Mustanski and Starks.

@@

More recently, a new statistical approach inspired from research in psychopathology has been used to model syndemics : network analysis [@Lee; @lee2020]. The core principle of this approach is that mental disorders arise from causal interactions between symptoms in a network [@borsboom2017]. Applied to syndemic theory, a syndemic would then be the network and the various conditions would then be the nodes of the network. Furthermore, one of the benefits of this approach resides in the theoretical possibility to identify influential nodes in a network : nodes that play a significant part in the activation of the network and that could be the prime target of an intervention in order to deactivate the network [@robinaugh2016]. In short, in addition to being statistically sound and coherent with the model of mutually causal epidemics, network analysis could offer actionable data to improve the health of MSM by identifying which conditions should be prioritized for interventions [@Tsai2018a].

We recommend that future research should abandon the summation score approach and attempt to clarify which model of interaction receives the most empirical support in order to implement effective public health interventions. Indeed as reminded by Chakrapani et al. the three models have different programmatic implications [@Chakrapani2019]. In the case of synergistically interacting epidemics, a single intervention may yield greater health improvement than if no interaction was present. For mutually causal epidemics, multicomponent interventions addressing the various epidemics should be implemented in order to produce an improvement. Finally, for serially causal epidemics, intervening at the root cause of a syndemic may prevent the development of a syndemic cascade.

@@

In qualitative studies, resilience was also found to exert a protective effect against syndemic conditions [@Chakrapani2019a; @Reed2016; @adam2018]. Furthermore, the review by Woodwards et al. found a protective effect of most resilience resources identified against HIV [@woodward2017].

@@

Minority stress represents another interesting framework as it allows to explain how social stigma can get under the skin of minorized populations [@Meyer2003].

@@

Some scholars are skeptical on the usefulness of syndemic theory to intersectionality and question the possibility of a conversation between the two fields [@sangaramoorthy2021]. As pertinent as the authors commentary is, we do believe in the possibility of making the two field converse. As Quinn pointed out in her response, incorporating an intersectional framework into syndemic research may address issues of power, oppression and structural violence, neglected thus far in the field [@quinn2021]. Doing so, we may attend to some gap in the literature and disentangle the complex web of stigma and structural violence faced by sexual minority men, especially MSM situated at the intersection of multiple form of systemic discrimination such as racism, homophobia, biphobia, transphobia, classism, ableism or stigma against sex workers or people living with HIV.
